---
title: "iRODS new HTTP API 🥳"
subtitle: "Release 0.2.0 of rirods"
author: 
  - "Martin Schobben"
  - "Mariana Montes"
  - "Terrell Russell"
date: "08/07/2023"
date-format: long
categories: [news, R, FAIR, open science]
---

*This article is about release 0.2.0 of rirods (see the [documentation](http://rirods.irods4r.org)).*


```{r message=FALSE}
library(readr)
library(dplyr)
library(forcats)
library(ggplot2)

commits <- read_csv("https://raw.githubusercontent.com/FAIReLABS/iRODS4R/main/posts/rirods-0-1-1/data/irods_commits.csv")


.old_config_dir <- Sys.getenv("R_USER_CONFIG_DIR")
.old_wd <- setwd(tempdir())
Sys.setenv("R_USER_CONFIG_DIR" = tempdir())

commits <- commits |> 
  filter(!fork, !archived, !private, !is.na(commits_names)) |> 
  mutate(
    commits_names = fct_lump(commits_names, 10),
    commits_names = fct_infreq(commits_names)
  )

ggplot(commits,aes(x = commits_names)) + 
    geom_bar() + 
    coord_flip()
```


Here comes `rirods` into the picture. Install the package from CRAN or get the development version from [GitHub](https://github.com/irods/irods_client_library_rirods).

```{r eval=FALSE}
install.packages("rirods")
```

We will demonstrate `rirods` usage by running a mock server on your localhost. The username and password are set to `"martin"` and `"secret"` for the example, but can be changed accordingly.

*Note that one needs Bash and Docker (with docker-compose) to use this mock-up.*

```{r}
library(rirods)
use_irods_demo("martin", "secret")
```

We follow the instructions from the output and create a project file with the server details and login to the system with our credentials.

```{r}
create_irods("http://localhost:9001/irods-http-api/0.2.0")
iauth("martin", "secret")
```

Now we can save the R object `commits` with `isaveRDS()`, which behaves similar to base R's `saveRDS()`.

```{r}
isaveRDS(commits, "commits.rds")
```


```{r}
setwd(.old_wd)
Sys.setenv("R_USER_CONFIG_DIR" = .old_config_dir)
```

